Diffusion of Innovation in the Open Source Software Ecosystem

### 2014-08-25 Paper/paper.tex
Many minor changes Paper is nearly ready, submitting...

\begin{abstract}
Innovative open source software projects have been at the centre of innovation in the software industry. However, the dynamics of open source project innovation and adoption remain under-researched.
In this paper we make the case that software package dependencies are a form of  adoption of innovation. We then test this on the time-evolution of the Gentoo package dependency graph. We find that the Bass model of innovation diffusion fits the growth of the number of packages depending on a given library with an $\overline{R}^2 > 98\%$.
\end{abstract}

@@

In this paper we not only analyse the adoption of open source libraries by other projects, but also use a method to test the diffusion of innovation of a software project - or the lifecycle of a project's adoption. We analyse the adoption of the open source libraries using the theoretical lens of \emph{diffusion of innovation} \citep{rogers03} and test it with a particular formalisation borrowed from the field of marketing research namely the \emph{Bass diffusion model} \citep{bass69}. Our central hypothesis is that  \emph{the growth in the number of open source software dependees of a project (its adoption) is a Bass innovation diffusion process.} In particular, we study  software innovation of the libraries in the Gentoo Portage  database from an adoption perspective. We first download the Gentoo Portage database for a time line of one month over a period of around 13 years (from 1 January to around mid 2012). We then calculate the dependencies among the libraries in each version of the database. Then, we calculate the change in the dependency graph of some well known libraries and check if this adoption of the library fits the Bass diffusion model \citep{bass69}.  We then analyse the adoption of 10 randomly selected libraries (we discard and replace those libraries whose adoption trends contain a lot of noise due to various reasons) compared to the Bass diffusion model. In all cases we find that the Bass diffusion model explains the adoption of the libraries - with an $\overline{R}^2 > 98\%$. This finding shows that the adoption of open source software libraries (in the Gentoo Portage database) does not differ from the adoption of other kinds of innovations as predicted by the theory of adoption of innovation, and in particular the Bass diffusion model. This finding, if found generalisable to most large open source libraries, can help in predicting their adoption over their lifetime. This can hence provide guidance to package managers and users of open source libraries.

@@

In the next section we describe the Bass \citep{bass69}  diffusion model, followed by the results of our analysis of fitting the Bass diffusion model onto the change in the adoption of the packages belonging to the Gentoo portage database.

@@

We found the Gentoo Portage data source to satisfy all these conditions, especially with regard to the minting the project dependencies and the frequent release cycles (also see Appendix B).
Now that a data source is selected, we extract the required information and process the data into a form that allows easy calculations. The major steps are collecting the raw data, parsing this into a simpler format and producing the final dependency graph from this simpler form. Some post-processing can then be done on the dependency graph. In the process, the dataset we find that the dataset shrunk from thirteen gigabytes (taking more than a week to collect), to thirty megabytes that could be processed in four seconds.

@@

% Package provides
Using all the observations and choices made above, a tool was developed to extract dependency graphs from all the simplified database snapshots. The shear scale of the resulting dataset, $1.3$ million packages and $6.9$ million dependency relations, required a solution to store efficiently. Therefore, a compressed format was developed that only stores the changes between the historical snapshots instead of storing whole snapshots. The extraction process took three hours and resulted in a $29$ megabyte file. Reading this file into memory resident data structures and performing simple queries took about $4$ seconds. The dataset is now in a usable form. The dataset and the details of the data structure can be seen in Appendix 1.

@@

Special tools where developed to extract the time series of the number of adopters $A_t$ for a given package. This time series was then fit to \eqref{eq:bass} using Mathematica's \texttt{NonLinearModelFit}. The goodness-of-fit was analysed using an ANOVA table and calculated using the adjusted coefficient of determination $\overline{R}^2$. The parameters were extracted from the fit, and confidence intervals were calculated under assumptions of normality. In table \ref{tbl:results} the relevant parameters are presented with their mean value and a $95\%$ confidence interval. Since normality was assumed, the confidence intervals ignore the $p \ge 0$ and $q \ge 0$ constraints.

The plots were drawn using a red line for the model and shades of red for the confidence bands. The red line was drawn according to \eqref{eq:bass}, with the mean values used as parameters. Then single value predictions confidence bands were calculated for $90\%$, $95\%$, $99\%$ and $99.9\%$ confidence levels and these were drawn in progressively darker shades of pink. They represent a prediction for where a single additional value would likely fall. According to the model and the uncertainty introduced by the fit, there is a $90\%$ chance that it will fall in the innermost band, $95\%$ chance that it would fail in the second band, etc. If the data fits the model properly, one would expect to see $90\%$ of the points in the inner band. Finally, the empirical data points were plotted as black dots, connected with a thin vertical line - the residual error of the model.

@@

The first package we consider is \verb|git|, a modern revision control system that shows an imitator driven adoption. It's growth can be seen in figure \ref{fig:git}, the corresponding statistics are in table \ref{tbl:results}. The package first appeared just before 2005, it had around ten packages depending on it in 2006, twenty in 2008 and is currently used by almost three hundred packages. According to the Bass model, it will continue to grow to approximately 750 users. The innovator inflow is only $0.2\%$ of the potential market per year, so one would expect $0.002\cdot (750 - 300) = 1$ user to adopt \verb|git| out of sheer innovation. Taking the analogy with persons, if someone from the $450$ current non-users where to meet a random person from the entire $750$ market, there is a $\frac{300}{750}= 40\%$ chance of meeting a user which can convince him/her to start using \verb|git|. The chance of this happening is the imitator inflow $q = 0.73$. Therefore, the total number of users \verb|git| can expect to gain from imitation this year is $450\cdot 40\%\cdot 0.73 = 131$. Very much imitator driven!

The relative slowness of \verb|git|'s growth and its dependence on imitator can be explained. Open source projects, and software project in general, consist of numerous large textual files containing source code. Changes made in one place can hugely and unpredictably affect other places. To complicate matters further, usually more than one developer works on the source code at the same time. There are competing systems such as \verb|cvs|, \verb|subversion|, \verb|mercurial|, etcetera., but the basic functionality of maintaining version is provided by all of them. Thus two explanations can be derived for \verb|git|'s growth: First, the revision control system is not a part that affects the products delivered by the open source project and second, there is little incentive to switch unless the new revision control system is proved to be superior.

@@

\verb|cairo| is a graphics library. It provides facilities for drawing lines, circles, text and other graphics primitives and is used by user graphics-heavy projects such as user interface libraries. Much like \verb|udev| it is an architectural change at a low level, this might explain its similar growth pattern.

@@

A typical example of innovator driven growth is given by \verb|libmad|. The model is fitted resulting in figure \ref{fig:libmad}. Again, the data is neatly explained by a Bass diffusion process, in particular the rapid steep growth and the stable user base afterwards. The name is an acronym for ``library for MPEG Audio Decoding'' and the package provides a high quality mp3 decoder for use in multimedia applications. This might also explain the rapid growth of its adoption: multimedia applications can benefit a lot from good quality mp3 support.

@@

A Bass diffusion increases monotonically, and never declines. However, not all packages follow this behaviour. Project \verb|libmad| (see \ref{fig:libmad}) is a good example of this contra-behaviour. The package has an innovator driver growth that brings it close to its maximum in about two years. After that, the package's usage remains almost flat for years, and will do so indefinitely if it is a perfect Bass diffusion process. This is called the ``maturity stage'' in product life-cycle parlance.

A real product life-cycle will also include a ``decline stage'' where the product begins to become obsolete. The Bass innovation diffusion model does not account for this. In a deep sense it would not have to, once ideas spread they become part of our collective knowledge and will continue to be used by the new products being developed. But the Bass model was not developed for the spreading of ideas, it was developed in the context of marketing to model the adoption of products. Extending the Bass model to include obsolescence would be an interesting extension for future research.

The package \verb|xulrunner| in the dataset is a nice example of a short but complete life cycle, see \ref{fig:xulrunner}. When the Bass model is applied naively and a least mean squares best-fit is made, the result is a poor fit, with $\overline{R}^2 = 96.8\%$. If one looks at the dependency growth of the package, the cause is clear: the package becomes obsolete, which the Bass model as presented in section \ref{eq:bass} does not represent. The decline of the package from approximately 2011 onwards can be seen as blue dots in the figure.

@@

The above examples of Bass diffusion growth suggests the central hypotheses of this article: \emph{the growth in the number of open source software dependees of a project (its adoption) is a Bass innovation diffusion process.} This hypotheses is tested using a methodic approach: a random selection of ten software packages is taken and their dependee growth is fitted to the Bass diffusion model.

The Gentoo portage database contains over twenty thousand packages, but only a fraction of these packages are suitable. About six thousand packages where introduced before 2004, where the growth of Gentoo itself may be confounded. Of the remaining packages, the vast majority do not have a number of other packages depending on it, these would for example be applications (as opposed to libraries) or unpopular/niche libraries. These are filtered out by setting a lower bound for the maximum number of dependees a package has during its lifetime. Then there are infrastructural packages in Gentoo, that have no corresponding open source projects, but are used manage the package database. In summary:

@@

After this initial automated filtering, packages are randomly selected and manually tested against further selection criteria. If a package is rejected a new one is randomly selected and the process is repeated until ten packages have been found. The final criteria are:

\begin{enumerate}
	\item[A.] The dependee growth does not contain strong discontinuities. It is very uncommon that many independent software project suddenly, within the time resolution of one month, decide to start or stop using a particular library. It is however common that maintainers of Gentoo Portage do re-organisations or mass additions/removals of packages that produce such discontinuities. Since they do not reflect actual adoption they are noise and are filtered out.
	\item[B.] The package corresponds to an open source software project. Some of the packages represent closed source software projects or non-software projects. These are not part of the scope and are therefore excluded.
	\item[C.] The package has not been forked of or factored out of another package. Some libraries where initially part of a different library that got refactored and split up in separate libraries and packages. Since projects may have been depending on the initial library and simply switch over to the refactored one, this can not be considered adoption of the library. Such packages are therefore excluded.
	\item[D.] The first included version should be close to a '1.0' release. The Bass diffusion hypotheses is about the entire growth of an open source software project, it is therefore necessary to have dependee growth data from close to the beginning of the project. Package that had an initial public release long before inclusion in Gentoo Portage are therefore excluded.
\end{enumerate}

These criteria are verified by consulting the information available through url \eqref{url:pkg}, in particular: the \texttt{changelog} of the \texttt{ebuild} file, any \texttt{readme} or \texttt{changelog} included in the earliest available version of the source code, the projects revision history and the projects homepage.

The manually rejected packages are listed in table \ref{tbl:rejected}. It should be noted that the selection criteria are chosen strictly to ensure that the main hypotheses is tested correctly, with independent open source software projects taken from their introduction and in their whole. When looking at the rejects, it does appear that some of them (\texttt{app-text/asciidoc} and many of the \texttt{x11-libs/} packages) still follow the Bass diffusion model.

The \texttt{x11-libs/} libraries where disqualified because they are the result of a refactor. In 2004 the X ecosystem re-organised, resulting in a fork of the XFree86 codebase and a subsequent decision to split the project up in smaller libraries. The current version was maintained in minor releases in both a split and un-split variety, but with the next major version the un-split branch was dropped. The projects show a innovator driven rapid Bass diffusion. This can be explained as the adoption of the new split release by the pre-existing users of the un-split version. The steepness can be explained by the dependers not wanting to be left behind. The innovation based drive can be explained by the relative safety and simplicity of upgrading to a new major version. A similar library-split refactor happened with the Qt library, with similar results.

After rejecting the packages above we arrive at the our final sample of 10 open source packages in table \ref{tbl:selected}.

@@

The packages \texttt{dev-dotnet/lxml} and \texttt{dev-cvs/mercurial} show a staircase-shaped deviation from the Bass diffusion model. This can be explained as discretisation noise. The number of dependees is integer and can therefore only rise or fall in discrete steps, whereas the Bass model is continuous and smooth. This introduces additional variance which lowers the $\overline{R}^2$ metric. It also appears to happen in \texttt{dev-haskell/cabal}, but this is an illusion since the steps in the growth are not singular but in the order of tens of packages. The reasons for these steps are unknown and likely related to Gentoo internal matters. \texttt{dev-libs/libunique} does not fit the Bass diffusion model well, and it appears to be for the same reasons that have been presented earlier for \texttt{xulrunner}. Despite this, all packages show $\overline{R}^2$ values over $98\%$.

@@

The growth of the number of packages depending on a packages can be modelled as a a Bass diffusion process. Overall the Bass diffusion model gave very a good fit for most OSS projects. Using only four parameters, it was able to describe the growth curves from the empirical data. Full statistical rigour would require a more involved analysis using the methods from, for example, \citet{carlos06}, but given the amount of and quality of evidence found we can conclude that most OSS project do follow the Bass diffusion model.

@@

Analysing the package dependency graph and its changes over time can provide new insights. Our exploratory study provides some evidence for insights such as how multimedia libraries are being adopted through an innovator driven process with low-level architectural changes happening slowly and through imitation. Further studies could test these hypotheses.

The package dependency graph contains empirical data to test extensions of the Bass diffusion model - extended with discarders. The Bass model and the present analysis is formulated in terms of absolute number of users, but in most applications only sales figures are available. The amount of sales is the first derivative of the Bass model, hence the model is usually applied in its derivative form \citep{mahajan95}. As a consequence the model only considers adopters, but does not consider \emph{discarders}. In the \texttt{xulrunner} example, we see the package being discarded from 2011 onwards, providing insights in the discarding mechanism. The next step would be to collect more examples of packages being discard, look at their patterns of demise and develop a model of discarding to supplement the Bass model of adoption. One model could for example be the inverse of a Bass curve, this makes sense when the market share of the original package is taken over by a new package. The unique feature of dependency graph analysis to give absolute user numbers facilitates this. Such an analysis can also help in predicting the behaviour of certain projects - hence forewarning the particular project stakeholders. As the extent of adoption of an open source project is an indication of how popular the project is and in-turn an indication of the health of the project as assessed by the open source community, we can typically supplement the existing analysis of open source project health \citep{daniel13,amrit10a,piggot13}.

The scale and complexity of the dependency graphs and open source innovation requires some care. Three notable issues became apparent in this study: First, In the open source community there is a lot of forking. It is not always clear whether a forked project constitutes the continuation of the original project or a separate new project. A more thorough study on the nature of forking could provide the insights to resolve this. Second, due to the public nature of open source development many immature or abandoned projects are visible in the larger datasets. This is good from a scientific perspective: it allows one to research projects from their early beginning and look at projects that failed to grow or became obsolete. But it clouds the `big picture' with many projects that do not significantly contribute to the overall innovation. In large datasets one would have to devise a relevance metric to select the relevant metrics. Such metrics could be the number of developers, the number of users or the number of dependees. Third, the sheer scale of the available OSS databases provide challenges for analysis. Specialist tooling is required to transform the raw data into more manageable formats.

@@

Still, there can be a bias in the selection of packages. Ubuntu, for example, has a specific target audience of novice Linux users and provides them with a Gnome based desktop environment. This means there is less pressure incentive for Ubuntu to include other desktop environments and advanced server software in their package database. Despite this, Ubuntu's database still includes a few other desktop environments and various advanced packages. Broad support appears more important than the burden of supporting many components. In general, the distributions have a user base that is very broad, some distributions are used from large server farms to mobile device and provides software packages for users that range from children to professionals in various fields.

A distinctive advantage of distribution package databases is that they contain very good dependency information. As mentioned earlier the package database needs to contain correct dependency for correct operation of the system and since the package manager operates automatically this information needs to be fully codified. With relative ease one can extract the entire dependency graph from the package database. This greatly facilitates gathering a dataset.

Source distributions will necessarily contain more a more complete dependency graph than binary distributions. Both kinds of package databases need to list all the dependencies required to \emph{use} a particular package, but source base distribution need to include all de packages required to \emph{compile} and \emph{use} the project.

The Gentoo portage database and the FreeBSD ports databases are ideal candidates. Both have all the good qualities of a source based package database. FreeBSD's has the advantage that it has a long history and more packages, Gentoo's has the advantage that it has better time resolution. Since the first author is very familiar with Gentoo's database, this one is chosen.

>>>>

\begin{abstract}
Open source software projects have been at the centre of innovation in the software industry. However, the dynamics of open source project innovation and adoption remain under-researched. This is especially the case in large open source software ecosystems, where projects can depend on many projects as well as have many dependers.

In this paper we make the case that software package dependencies are a form of  adoption of innovation. We then test this on the time-evolution of the Gentoo package dependency graph. We find that the Bass model of innovation diffusion fits the growth of the number of packages depending on a given library with an $\overline{R}^2 > 98\%$.
\end{abstract}

@@

In this paper we not only analyse the adoption of open source libraries by other projects, but also use a method to test the diffusion of innovation of a software project - or the lifecycle of a project's adoption. We analyse the adoption of the open source libraries using the theoretical lens of \emph{diffusion of innovation} \citep{rogers03} and test it with a particular formalisation borrowed from the field of marketing research namely the \emph{Bass diffusion model} \citep{bass69}. Our central hypothesis is that  \emph{the growth in the number of open source software dependees of a project (its adoption) is a Bass innovation diffusion process.} In particular, we study  software innovation of the libraries in the Gentoo Portage  database from an adoption perspective. We first download the Gentoo Portage database for a time line of one month over a period of around 13 years (from 1 January to around mid 2012). We then calculate the dependencies among the libraries in each version of the database. Then, we calculate the change in the dependency graph of some well known libraries and check if this adoption of the library fits the Bass diffusion model \citep{bass69}.  We then analyse the adoption of 10 randomly selected libraries (we discard and replace those libraries whose adoption trends contain a lot of noise due to various reasons) compared to the Bass diffusion model. In all cases we find that the Bass diffusion model explains the adoption of the libraries (with an $\overline{R}^2 > 98\%$). This finding shows that the adoption of open source software libraries (in the Gentoo Portage database) does not differ from the adoption of other kinds of innovations as predicted by the theory of adoption of innovation, and in particular the Bass diffusion model. This finding, if found generalisable to most large open source libraries, can help in predicting their adoption over their lifetime. This can hence provide guidance to package managers and users of open source libraries.

@@

In the next section we describe the Bass diffusion model \citep{bass69}, followed by the results of our analysis of fitting the Bass diffusion model onto the change in the adoption of the packages belonging to the Gentoo portage database.

@@

We found the Gentoo Portage data source to satisfy all these conditions, especially with regard to having more complete project dependencies and frequent release cycles (also see Appendix B).
Now that a data source is selected, we extract the required information and process the data into a form that allows easy calculations. The major steps are collecting the raw data, parsing this into a simpler format and producing the final dependency graph from this simpler form. Some post-processing can then be done on the dependency graph. In the process, the dataset we find that the dataset shrunk from thirteen gigabytes (taking more than a week to collect), to thirty megabytes that could be processed in four seconds.

@@

% Package provides
Using all the observations and choices made above, a tool was developed to extract dependency graphs from all the simplified database snapshots. The shear scale of the resulting dataset, $1.3$ million packages and $6.9$ million dependency relations, required a solution to store efficiently. Therefore, a compressed format was developed that only stores the changes between the historical snapshots instead of storing whole snapshots. The extraction process took three hours and resulted in a $29$ megabyte file. Reading this file into memory resident data structures and performing simple queries took about $4$ seconds. The dataset is now in a usable form. Table \ref{tbl:dataset} provides the key figures of the dataset, while the details of the data structure can be seen in Appendix 1.

@@

Special tools where developed to extract the time series of the number of adopters $A_t$ for a given package. This time series was then fit to equation \eqref{eq:bass-model} using Mathematica's \texttt{NonLinearModelFit}. The goodness-of-fit was analysed using an ANOVA table and calculated using the adjusted coefficient of determination $\overline{R}^2$. The parameters were extracted from the fit, and confidence intervals were calculated under assumptions of normality. In table \ref{tbl:results} the relevant parameters are presented with their mean value and a $95\%$ confidence interval. Since normality was assumed, the confidence intervals ignore the $p \ge 0$ and $q \ge 0$ constraints.

The plots were drawn using a red line for the model and shades of red for the confidence bands. The red line was drawn according to equation \eqref{eq:bass-model}, with the mean values used as parameters. Then single value predictions confidence bands were calculated for $90\%$, $95\%$, $99\%$ and $99.9\%$ confidence levels and these were drawn in progressively darker shades of pink. They represent a prediction for where a single additional value would likely fall. According to the model and the uncertainty introduced by the fit, there is a $90\%$ chance that it will fall in the innermost band, $95\%$ chance that it would fail in the second band, etc. If the data fits the model properly, one would expect to see $90\%$ of the points in the inner band. Finally, the empirical data points were plotted as black dots, connected with a thin vertical line - the residual error of the model.

@@

The first package we consider is \verb|git|, a modern revision control system that shows an imitator driven adoption. It's growth can be seen in figure \ref{fig:git}, the corresponding statistics are in table \ref{tbl:results}. The package first appeared just before 2005, it had around ten packages depending on it in 2006, twenty in 2008 and is currently used by almost three hundred packages. According to the Bass model, it will continue to grow to approximately 750 users. The innovator inflow is only $0.2\%$ of the potential market per year, so one would expect $0.002\cdot (750 - 300) = 1$ users to adopt \verb|git| out of sheer innovation. Taking the analogy to people, if someone from the $450$ current non-users were to meet a random person from the entire $750$ market, there is a $\frac{300}{750}= 40\%$ chance of meeting a user which can convince him/her to start using \verb|git|. The chance of this happening is the imitator inflow $q = 0.73$. Therefore, the total number of users \verb|git| can expect to gain from imitation this year is $450\cdot 40\%\cdot 0.73 = 131$. Implying that the market is very much imitator driven.

The relative slowness of \verb|git|'s growth and its dependence on imitation can be explained. Open source projects, and software projects in general, consist of numerous large textual files containing source code. Changes made in one place can hugely and unpredictably affect other places. To complicate matters further, usually more than one developer works on the source code at the same time. There are competing systems such as \verb|cvs|, \verb|subversion|, \verb|mercurial|, etc., but the basic functionality of maintaining versions is provided by all of them. Thus two explanations can be derived for \verb|git|'s growth: first, the revision control system is not a part that affects the products delivered by the open source project and second, there is little incentive to switch unless the new revision control system proves to be superior.

@@

\verb|cairo| is a graphics library. It provides facilities for drawing lines, circles, text and other graphics primitives and is used by user graphics-heavy projects such as user interface libraries. Much like \verb|udev| its adoption represents an architectural change at a low level, this might explain its similar growth pattern.

@@

A typical example of innovator driven growth is given by \verb|libmad|. The model is fitted resulting in figure \ref{fig:libmad}. Again, the data is neatly explained by a Bass diffusion process and in particular, the rapid steep growth and the stable user base afterwards. The name of the project is an acronym for ``library for MPEG Audio Decoding'' and the package provides a high quality mp3 decoder for use in multimedia applications. This might also explain the rapid growth of its adoption: multimedia applications can benefit a lot from good quality mp3 support.

@@

A Bass diffusion increases monotonically, and never declines. However, not all packages follow this behaviour. Project \verb|libmad| (see Figure \ref{fig:libmad}) is a good example of this contra-behaviour. The package has an innovator driver growth that brings it close to its maximum in about two years. After that, the package's usage remains almost flat for years, and will do so indefinitely, if it is a perfect Bass diffusion process. This is called the ``maturity stage'' in product life-cycle parlance.

A real product life-cycle will also include a ``decline stage'' where the product begins to become obsolete. The Bass innovation diffusion model does not account for this. In a deep sense it would not have to as once ideas spread they become part of our collective knowledge and will continue to be used by the new products being developed. But the Bass model was not developed for the spreading of ideas, it was developed in the context of marketing, to model the adoption of products. Extending the Bass model to include obsolescence would be an interesting extension for future research.

The package \verb|xulrunner| in the dataset is a nice example of a short but complete life cycle (see Figure \ref{fig:xulrunner}). When the Bass model is applied naively and a least mean squares best-fit is made, the result is a poor fit, with $\overline{R}^2 = 96.8\%$. If one looks at the dependency growth of the package, the cause is clear: the package becomes obsolete, which the Bass model as presented in equation \ref{eq:bass-model} does not represent. The decline of the package from approximately 2011 onwards can be seen as blue dots in the figure.

@@

The above examples of Bass diffusion growth suggests the central hypothesis of this article: \emph{the growth in the number of open source software dependees of a project (its adoption) is a Bass innovation diffusion process}. This hypothesis is tested using a methodic approach: a random selection of ten software packages is taken and their dependee growth is fitted to the Bass diffusion model.

The Gentoo portage database contains over twenty thousand packages, but only a fraction of these packages are suitable. About six thousand packages where introduced before 2004, where the growth of Gentoo itself may be confounded. Of the remaining packages, the vast majority do not have a number of other packages depending on it, these would for example be applications (as opposed to libraries) or unpopular/niche libraries. These are filtered out by setting a lower bound for the maximum number of dependees a package has during its lifetime. Then there are infrastructural packages in Gentoo, that have no corresponding open source projects, but are used to manage the package database. In summary:

@@

After this initial automated filtering, packages are randomly selected and manually tested against further selection criteria. If a package is rejected, a new one is randomly selected and the process is repeated until ten packages have been found. Our selection criteria are:

\begin{enumerate}
	\item[A.] The dependee growth does not contain strong discontinuities. It is very uncommon that many independent software project suddenly, within the time resolution of one month, decide to start or stop using a particular library. It is however common that maintainers of Gentoo Portage do re-organisations or mass additions/removals of packages that produce such discontinuities. Since they do not reflect actual adoption they are noise and need to be filtered out.
	\item[B.] The package corresponds to an open source software project. Some of the packages represent closed source software projects or are not software projects. These are not part of the scope and are therefore excluded.
	\item[C.] The package has not been forked of or factored out of another package. Some libraries were initially part of a different library that got refactored and split up in separate libraries and packages. Since projects may have been depending on the initial library and simply switch over to the refactored one, this cannot be considered adoption of the library. Such packages are therefore excluded.
	\item[D.] The first included version should be close to a '1.0' release. The Bass diffusion hypothesis is about the entire growth of an open source software project, it is therefore necessary to have dependee growth data from close to the beginning of the project. Packages that had an initial public release long before inclusion in Gentoo Portage are therefore excluded.
\end{enumerate}

These criteria are verified by consulting the information available through url \eqref{url:pkg}, in particular: the \texttt{changelog} of the \texttt{ebuild} file, any \texttt{readme} or \texttt{changelog} included in the earliest available version of the source code, the project's revision history and the project's homepage.

The manually rejected packages are listed in table \ref{tbl:rejected}. It should be noted that the selection criteria are chosen strictly to ensure that the main hypothesis is tested correctly, with independent open source software projects taken from their introduction and in their whole. When looking at the rejects, it does appear that some of the (\texttt{app-text/asciidoc} packages and many of the \texttt{x11-libs/} packages) still follow the Bass diffusion model.

The \texttt{x11-libs/} libraries did not qualify because they are the result of a refactor. In 2004 the X ecosystem re-organised, resulting in a fork of the XFree86 codebase and a subsequent decision to split the project up in smaller libraries. The current version was maintained in minor releases in both a split and un-split variety, but with the next major version the un-split branch was dropped. The projects show a innovator driven rapid Bass diffusion. This can be explained as the adoption of the new split release by the pre-existing users of the un-split version. The steepness can be explained by the dependers not wanting to be left behind. The innovation based drive can be explained by the relative safety and simplicity of upgrading to a new major version. A similar library-split refactor happened with the Qt library, with similar results. The complete list of the rejected packages can be seen in Table \ref{tbl:rejected}.

After rejecting the packages shown in Table \ref{tbl:rejected}, we arrive at the our final sample of 10 open source packages in Table \ref{tbl:selected}.

@@

The packages \texttt{dev-python/lxml} (Figure \ref{fig:fit77}) and \texttt{dev-cvs/mercurial} (Figure \ref{fig:fit242}) show a staircase-shaped deviation from the Bass diffusion model. This can be explained as discretisation noise. The number of dependees is an integer and can therefore only rise or fall in discrete steps, whereas the Bass model is continuous and smooth. This introduces additional variance which lowers the $\overline{R}^2$ metric. It also appears to happen in \texttt{dev-haskell/cabal} (Figure \ref{fig:fit65}), but this is an illusion since the steps in the growth are not singular but in the order of tens of packages. The reasons for these steps are unknown and likely related to Gentoo internal matters. \texttt{dev-libs/libunique} (Figure \ref{fig:fit218}) does not fit the Bass diffusion model well, and it appears to be for the same reasons that have been presented earlier for \texttt{xulrunner}, i.e. that its adoption represents initial growth and later demise. Despite this, all packages show $\overline{R}^2$ values over $98\%$, as seen in Table \ref{tbl:testresults}.

@@

We have shown that the growth of the number of packages depending on a package can be modelled as a a Bass diffusion process \citep{bass69}. The Bass diffusion model is a theory that explains and predicts (also known as a type IV theory, according to \citep{gregor06}). What we have shown in this paper is a fit of this theory to explain the process by which open source libraries are adopted. Overall we see that the Bass diffusion model gives very a good fit for most OSS projects. Using only four parameters, it is able to describe the growth curves from the empirical data. Full statistical rigour would require a more involved analysis using the methods from, for example, \citet{carlos06}. But given the amount of and quality of evidence found, we can conclude that most OSS project do follow the Bass diffusion model.

@@

Analysing the package dependency graph and its changes over time can provide new insights. Our exploratory study provides some evidence for insights such as how multimedia libraries are being adopted through an innovator driven process with low-level architectural changes happening slowly and through imitation. Further studies could test this and related hypotheses.

The package dependency graph contains empirical data to test extensions of the Bass diffusion model - extended with discarders. The Bass model and the present analysis is formulated in terms of absolute number of users, but in most applications only sales figures are available. The amount of sales is the first derivative of the Bass model, hence the model is usually applied in its derivative form \citep{mahajan95}. As a consequence, the model only considers adopters, but does not consider \emph{discarders}. In the \texttt{xulrunner} example, we see the package being discarded from 2011 onwards, providing insights into the discarding mechanism. The next step would be to collect more examples of packages being discarded, and looking at the patterns of demise and perhaps developing a model of discarding to supplement the Bass model of adoption. One model could for example be the inverse of a Bass curve, this makes sense when the market share of the original package is taken over by a new package. The unique feature of dependency graph analysis to give absolute user numbers facilitates this. Such an analysis can also help in predicting the behaviour of certain projects - hence forewarning the particular project stakeholders. As the extent of adoption of an open source project is an indication of how popular the project is and in-turn an indication of the health of the project as assessed by the open source community, we can use this predictive adoption to typically supplement the existing analysis of open source project health \citep{daniel13,amrit10a,piggot13}.

The scale and complexity of the dependency graphs and open source innovation requires some attention. Three notable issues became apparent in this study: first, that in the open source community there is a lot of forking. It is not always clear whether a forked project constitutes the continuation of the original project or a separate new project. A more thorough study on the nature of forking could provide the insights to resolve this issue. Second, due to the public nature of open source development many immature or abandoned projects are visible in the larger datasets. This is good from a scientific perspective as it allows one to research projects from their beginning stages and one can look at projects that failed to grow or those that became obsolete. But it clouds the `big picture' with many projects that do not significantly contribute to the overall innovation. In large datasets one would have to devise a relevance metric to select the relevant metrics. Such metrics could be the number of developers, the number of users or the number of dependees. Third, the sheer scale of the available OSS databases provide challenges for analysis. Specialist tooling is required to transform the raw data into more manageable formats.

@@

A distinctive advantage of distribution package databases is that they contain very good dependency information. The package database needs to contain correct dependency for correct operation of the system and since the package manager operates automatically this information needs to be fully codified. With relative ease one can extract the entire dependency graph from the package database. This greatly facilitates gathering a dataset.

\emph{Source} distributions will necessarily contain more a more complete dependency graph than \emph{binary} distributions. Both kinds of package databases need to list all the dependencies required to \emph{use} a particular package, but source base distribution need to include all de packages required to \emph{compile} and \emph{use} the project.

The Gentoo portage database and the FreeBSD ports databases are ideal candidates. Both have all the good qualities of a source based package database. FreeBSD has the advantage that it has a long history and more packages, Gentoo has the advantage that it has better time resolution. Also since the first author is very familiar with Gentoo's database, we chose the Gentoo portage database as the data source for this article.

For more details on the process of selecting the database please see \citep{bloemen12}.
